# Chapter 6. Analysis of longitudinal data

```{r}
library(patchwork)
library(tidyverse)
```

## 1. Read the data.

Read the data from data wrangling exercise and re-factor categorical variables.

```{r}
BPRS_long <- read.csv("data/BPRS_long.csv")
RATS_long <- read.csv("data/RATS_long.csv")

BPRS_long <- BPRS_long %>% 
  dplyr::mutate(subject = as.factor(subject),
                treatment = as.factor(treatment),
                week = as.factor(week))

RATS_long <- RATS_long %>% 
  dplyr::mutate(ID = as.factor(ID),
                Group = as.factor(Group),
                week = as.factor(week))
```

Visualize the response patterns of each respondent.

```{r}
g1 <- ggplot(BPRS_long, aes(x = week, y = value, group=subject, color=subject)) +
  geom_line() +
  facet_wrap(~treatment) +
  xlab("Week") +
  ylab("Symptoms") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.title.x = element_blank()
        ) 
g2 <- ggplot(BPRS_long, aes(x = week, y = value)) +
  geom_boxplot() +
  facet_wrap(~treatment) +
  xlab("Week") +
  ylab("Symptoms") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle=90))

g1 / g2
```
And for the RATS-data:

```{r}
g1 <- ggplot(RATS_long, aes(x = week, y = value, group=ID, color=ID)) +
  geom_line() +
  facet_wrap(~Group) +
  xlab("Week") +
  ylab("Symptoms") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.title.x = element_blank()
        ) 
g2 <- ggplot(RATS_long, aes(x = week, y = value)) +
  geom_boxplot() +
  facet_wrap(~Group) +
  xlab("Week") +
  ylab("Symptoms") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle=90))

g1 / g2
```

Standardize the datasets so that each weeks values are z-transformed independently from each other.

```{r}
BPRS_long_z <- BPRS_long %>%
  group_by(week) %>%
  mutate(value = scale(value)) %>%
  ungroup()

RATS_long_z <- RATS_long %>%
  group_by(week) %>%
  mutate(value = scale(value)) %>%
  ungroup()
```

Visualization for standardized data.

```{r}
g1 <- ggplot(BPRS_long_z, aes(x = week, y = value, group=subject, color=subject)) +
  geom_line() +
  facet_wrap(~treatment) +
  xlab("Week") +
  ylab("Symptoms") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.title.x = element_blank()
        ) 
g2 <- ggplot(BPRS_long_z, aes(x = week, y = value)) +
  geom_boxplot() +
  facet_wrap(~treatment) +
  xlab("Week") +
  ylab("Symptoms") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle=90))

g1 / g2
```

```{r}
g1 <- ggplot(RATS_long_z, aes(x = week, y = value, group=ID, color=ID)) +
  geom_line() +
  facet_wrap(~Group) +
  xlab("Week") +
  ylab("Symptoms") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.title.x = element_blank()
        ) 
g2 <- ggplot(RATS_long_z, aes(x = week, y = value)) +
  geom_boxplot() +
  facet_wrap(~Group) +
  xlab("Week") +
  ylab("Symptoms") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle=90))

g1 / g2
```

## Summary method

Let's produce summary tables from original datasets.

```{r}
BPRSS <- BPRS_long %>%
  group_by(treatment, week) %>%
  summarise(mean = mean(value),
            sd = sd(value),
            n = length(value),
            se = sd / sqrt(n) ) %>%
  ungroup()

RATSS <- RATS_long %>%
  group_by(Group, week) %>%
  summarise(mean = mean(value),
            sd = sd(value),
            n = length(value),
            se = sd / sqrt(n) ) %>%
  ungroup()
```
Summary graph for BPRS-data.

```{r}
ggplot(BPRSS, aes(x = week, y = mean, linetype = treatment, shape = treatment)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(bprs) +/- se(bprs)")
```

Summary graph for RATS-data

```{r, warning=FALSE, message=FALSE}
ggplot(RATSS, aes(x = week, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2, 3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8, 0.8)) +
  scale_y_continuous(name = "mean(rats) +/- se(rats)")
```